\chapter{Deterministic Mapping for Morphosyntax} % Main chapter title
\label{Chapter4}

The previous chapter talked about how CHAT and CoNLL-U use \texttt{MOR} and \texttt{GRA} codes to structurally represent morphosyntactic information. However, the set of tags used in the two annotation schemes could also be semantically different. This chapter presents a deterministic mapping from CHAT codes to UD-style labels. The mapping consists of three major categories: part-of-speech tags, morphological features and dependency relations and is created for English, French and Italian.

% converting MOR codes to UD Universal POS tags
\section{Part-of-Speech Tags}
\label{sec:pos}
% left-aligned table columns with automated line breaks
\newcolumntype{L}{>{\RaggedRight\arraybackslash}X}
\begin{margintable}[1\baselineskip] % move figure down by 1 line
\begin{tabularx}{\textwidth}{@{}Ll@{}}
\toprule
\textbf{MOR} & \textbf{UD}\\ \midrule
adj & ADJ\\
adv & ADV\\
coord & CCONJ\\
n & NOUN\\
num & NUM\\
pro & PRON\\
n:prop & PROPN\\
comp & SCONJ\\
v & VERB\\
det, qn & DET\\
co, fil & INTJ\\
dia, none & X\\
post, prep & ADP\\
aux, cop, mod & AUX\\
beg, end & PUNCT\\
inf, neg, part & PART\\\bottomrule
\end{tabularx}
\caption{\label{tab:martabpos}Examples of MOR POS categories that can be directly converted to UPOS tags. See Table \ref{tab:posmap1} and Table \ref{tab:posmap2} in \Cref{appendixa} for the full mapping.}
\footnotesize
% Examples of MOR POS categories that can be directly converted to UPOS tags. See \ref{tab:posmap1} and \ref{tab:posmap2} in \Cref{appendixa} for the full mapping.
\end{margintable}


The CHAT manual listed 39 POS tags for English (\cite{Macwhinney2000}, Part 3, p.8). This set can be extended with language-specific tags for other languages. UD, on the other hand, uses a universal POS (UPOS) tagset with only 17 tags for all languages in the UD project.

UD's UPOS is developed based on Petrov's \emph{A Universal Part-of-Speech Tagset}, where the crosslinguistic universality of the UPOS tags stems from their operational definitions instead of presupposing the existence of actual grammatical categories (\cite{petrov2012}). In CHAT, however, POS tags are much less universal across languages. Typologically different languages may use quite different tagsets than that prescribed by the MOR program for English (\cite{Macwhinney2000}, Part3, p.7). Additionally, researchers can define their own POS tagsets, which not only means that, in terms of POS tags, CHILDES took an opposite approach from UD, but also makes an exhaustive mapping between POS tags used in the two annotation schemes impossible. Therefore, the mapping I define here concerns only the standard POS tagset given by the MOR grammar for English, French and Italian.

The adaptation from MOR POS categories to UD's UPOS can be achieved in the following ways:
\begin{itemize}
	\item create a direct mapping between two tags if they share the same semantics or have a straightforward relationship
	\item collapse MOR POS categories into one UPOS tag when applicable
	\item make best-effort decisions based on the description of the POS cetegories by both formalisms
	\item map MOR POS categories with no UPOS counterparts to \texttt{X}
\end{itemize}

An example of a straightforward mapping between MOR POS codes and UPOS tags is shown in Table \ref{tab:martabpos}. A total of 16 out of the 17 UPOS tags are used in the tagset mapping, with the exception of \texttt{SYM}, which is reserved for symbols.

Collapsing of POS categories can be illustrated by the example of auxiliary verbs. The MOR code differentiates between auxiliary verbs \texttt{aux} or \texttt{v:aux}, copulas \texttt{cop} and modal verbs \texttt{mod}. However, since UD took an operational approach to define grammatical categories, it does not distinguish copulas or modal verbs from other verbs that function as auxiliaries. By doing so, UPOS can be applied to typologically distant languages.

The following MOR POS codes require more thoughts during the creation of a deterministic mapping, partially because they do not exist in all languages. For tags like these, UD usually raises them to a more coarse category and represents the exact type using morphological features instead.

\paragraph{\texttt{art}}
Articles are ganerally marked as \texttt{DET} in UD Treebanks with feature \texttt{PronType=Art} \sidenote{Definition for \texttt{DET} can be found from \url{https://universaldependencies.org/docs/u/pos/DET.html}}.

\paragraph{\texttt{preart}}
\texttt{preart} are contractions of prepositions and articles, for instance the Italian \emph{alla} is the contraction of \emph{a} and \emph{la}. In UD-style annotations, words like \emph{alla} are treated as multi-word tokens, and a POS tag is given to each part of the token. However, CHAT considers them to be single words with forms of their own. To keep the alignment with the \texttt{\%mor} tier, here \texttt{preart} is mapped to \texttt{DET} because the word functions as a determiner.

\paragraph{\texttt{vimp}}
Imperatives are tagged with \texttt{vimp} in \%mor tier. However, in UD, imperatives are tagged with \texttt{VERB}, which is the category for all finite verb forms \sidenote{Definition for \texttt{VERB} can be found from \url{https://universaldependencies.org/docs/u/pos/VERB.html}} . In UD, gender, person, number, tense, aspect, mood and voice are all specified as features. Therefore \texttt{vimp}, listed here as an example for finite verbs, is mapped to \texttt{VERB} with feature \texttt{MOOD=Imp}.

\paragraph{\texttt{vpfx}}
Preverb/verb particles are tagged with \texttt{vpfx}. Although they are particles, according to UD's definition \sidenote{Definition for \texttt{ADP} can be found from \url{https://universaldependencies.org/docs/u/pos/ADP.html}}, they belong to \texttt{ADP}.

\paragraph{\texttt{conj}}
Despite the similarity between the MOR \texttt{conj} and UD \texttt{CCONJ}, they represent different things. UD distinguishes between \texttt{CCONJ} \sidenote{Definition for \texttt{CCONJ} can be found from \url{https://universaldependencies.org/docs/u/pos/CCONJ.html}} and \texttt{SCONJ} \sidenote{Definition for \texttt{SCONJ} can be found from \url{https://universaldependencies.org/docs/u/pos/SCONJ.html}}, the former is for coordinating conjunctions like \emph{and} and \emph{but}, which are tagged \texttt{coord} by MOR, and the latter marks subordinating conjunctions with complementizers like \emph{that} or \emph{if}, or adverbial clause introducers like \emph{when}, these are tagged \texttt{conj} by MOR. Therefore, the suitable UPOS tag for the MOR POS tag \texttt{conj} is \texttt{SCONJ}.

\paragraph{special form markers}
It is worth noticing that some MOR codes do not correspond to any grammatical categories. For example, in CHAT, the so-called \emph{special form markers} have unique POS tags independent of the part-of-speech category to which the word belongs (\cite{Macwhinney2000}, Part 1, pp. 45-47). Nevertheless, some of these tags can still be mapped to UPOS.The tags \texttt{sing} for singing, \texttt{bab} for babbling and \texttt{wplay} for wordplay can all be mapped to \texttt{INTJ}, which stands for \emph{interjection}. The definition of interjection, however, is vague (\cite{ameka1992}). UD defines interjection descriptively \sidenote{"An interjection is a word that is used most often as an exclamation or part of an exclamation. It typically expresses an emotional reaction, is not syntactically related to other accompanying expressions, and may include a combination of sounds not otherwise found in the language." Excerpt from \url{https://universaldependencies.org/docs/u/pos/SCONJ.html}}. Following UD's definition for \texttt{INTJ}, the core criteria for a word \sidenote{or non-word in this case} to be an interjection include:
\begin{itemize}
	\item that it is often used exclamatorily
	\item that it typically expresses emotions
	\item that it plays no syntactic role in the utterance
	\item that it might not be a recognised word in the language
\end{itemize}
Since singing, babbling and wordplay satisfy these criteria, along with the MOR code \texttt{co} for interjections , they are granted the \texttt{INTJ} tag. Other special form markers, on the other hand, have no corresponding UPOS tags whatsoever, for example \texttt{dia} for dialect forms. Unfortunately they have to be assigned to \texttt{X}, as words that does not belong to the other 16 categories.

\begin{margintable}[1\baselineskip] % move figure down by 1 line
\begin{tabularx}{\textwidth}{@{}cl@{}}
\toprule
\textbf{Punctuations} & \textbf{MOR code}\\\midrule
‡ & beg\\
, & cm\\
„ & end\\
“ & bq\\
” & eq\\
‘ & bq2\\
’ & eq2\\\bottomrule
\end{tabularx}
\caption{\label{tab:martabpunct}MOR punctuation marks and their corresponding MOR codes.}
\footnotesize
\end{margintable}

\paragraph{punctuation marks}
Finally, MOR also defines a set of punctuation marks with string representations, shown in Table \ref{tab:martabpunct}. Although treated as lexical items in the \%mor tier, they are all given the \texttt{PUNCT} tag.

% feature mappings
\section{Morphological Features}
\label{sec:feats}

Since \texttt{MOR} uses a morpheme-based approach to analyse individual words, morphological features like \emph{tense} and \emph{mood} are encoded by morpheme codes. Unlike MOR's dissecting analysis of a word, UD takes a lexicalist view on morphology and treats word as a whole lexical item. Features are thus considered properties of the word. The existence of morphemes is not presupposed.

\begin{margintable}[1\baselineskip]
\begin{tabularx}{1\textwidth}{@{}ll@{}}
\toprule
\textbf{CHILDES database} & \textbf{Language}\\ \midrule
	Belfast & eng\\
	Brown & eng\\
	English-MiamiBiling & eng\\
	Wells & eng\\
	Geneva & fra\\
	Leveille & fra\\
	MTLN & fra\\
	vioncolas & fra\\
	Tonelli & ita\\\bottomrule
\end{tabularx}
\caption{\label{tab:martabdb}CHILDES databases chosen for this project.}
\footnotesize
\end{margintable}

These two fundamentally different views of words are reflected in the different ways of structuring morphological information. CoNLL-U format keeps all relevant features of a word in the token's \texttt{FEATS} field. And although each language in UD has its subset of morphological features, the feature types and values are by convention the same. Compared with UD, CHILDES's way of handling features is less direct. Not all features are explicit as UD's feature-value pairs. Instead, they have to be inferred from the morpheme codes or POS tags.

Listing of all possible MOR codes for grammatical morphemes is, again, unlikely. Therefore, I chose the easier path of collecting these codes from the CHILDES databases that I decided to work with, listed in \Cref{tab:martabdb}.
\begin{margintable}[1\baselineskip]
\begin{tabularx}{1\textwidth}{@{}ll@{}}
\toprule
\textbf{MOR} & \textbf{UD feats}\\ \midrule
	3s & Number=Sing|Person=3\\
	cp & Degree=Cmp\\
	f & Gender=Fem\\
	inf & VerbForm=Inf\\
	imp & Mood=Imp\\
	impf & Tense=Imp\\
	pass & Voice=Pass\\
	pastp & Tense=Past|VerbForm=Part\\
	poss & Poss=Yes\\\bottomrule
\end{tabularx}
\caption{\label{tab:martabfeats}Example MOR grammatical morpheme codes and their corresponding UD featue-value pairs.}
\footnotesize
\end{margintable}
Of all the morpheme codes found in these corpora, not all of them correspond to UD features. Since morphological features are not mandatory, I only created mappings for those with a corresponding UD feature-value pair. An example mapping is given in Table \ref{tab:martabfeats} on the right-hand side. Only 9 types of features can be obtained directly from the morpheme codes I encountered: \texttt{Number}, \texttt{Person}, \texttt{Gender}, \texttt{Mood}, \texttt{Degree}, \texttt{Tense}, \texttt{Verbform}, \texttt{Voice}, and \texttt{Poss}. All possible UD features and values allowed for Engish, French, and Italian can be found in Table \ref{tab:featsmap} in \Cref{appendixa}. The codes are sometimes capitalised and sometimes not due to language- or corpus-specific conventions. They are all converted to lowercase for easy comparison. The original morpheme codes are kept in the \texttt{MISC} field.
% \begin{margintable}[1\baselineskip]
% \begin{tabularx}{1\textwidth}{@{}ll@{}}
% \toprule
% \textbf{MOR code} & \textbf{UD feats}\\ \midrule
%     3s & Number=Sing|Person=3\\
%     cp & Degree=Cmp\\
%     f & Gender=Fem\\
%     inf & VerbForm=Inf\\
%     imp & Mood=Imp\\
%     impf & Tense=Imp\\
%     pass & Voice=Pass\\
%     pastp & Tense=Past|VerbForm=Part\\
%     poss & Poss=Yes\\\bottomrule
% \end{tabularx}
% \caption{\label{tab:martabfeats}Example MOR grammatical morpheme codes and their corresponding UD featue-value pairs.}
% \footnotesize
% \end{margintable}


% converting GRA codes to UD deprels
\section{Dependency Relations}
\label{sec:deprel}

In CHAT annotations, syntactic relations are represented by Grammatical Relations (GRs). They were first devised by Kenji Sagae (\cite{sagae-etal-2004-adding}) for the English language. After later revisions in 2007, CHAT now has 37 grammatical relation types (\cite{sagae2007}). They correspond to dependency relations stored in the \texttt{DEPRELS} field in CoNLL-U files. However, CHILDES and UD use different labels for these dependency relations and quite different parsing schemes. Sometimes, a conversion from GR to deprel also introduces a change in the structure of the dependency tree.

\begin{margintable}[1\baselineskip]
\begin{tabularx}{1\textwidth}{@{}Ll@{}}
\toprule
\textbf{GRA} & \textbf{UD deprel}\\ \midrule
SUBJ & nsubj\\
CSUBJ & csubj\\
OBJ & obj\\
OBJ2 & iobj\\
DET & det\\
APP & appos\\
AUX & aux\\
CONJ & conj\\
COORD & cc\\
PUNCT & punct\\
ROOT & root\\\bottomrule
\end{tabularx}
\caption{\label{tab:martabgr}Example CHILDES grammatical relations and their corresponding UD dependency relations.}
\footnotesize
\end{margintable}

Defining a deterministic mapping between GRs and UD dependency relations thus ranges from direct label translation to linguistically-informed decisions. \Cref{tab:martabgr} gives examples of direct label conversions. The full mapping for the current version of chatconllu can be found in \Cref{tab:grmap}.

 A solid mapping requires both familiarity with the grammatical relations defined in the two formalisms and profound linguistic knowledge for the languages involved, this mapping I created is only a first attempt with many inprecisions. Therefore, instead of explaining the decision process with my limited linguistic knowledge, I choose to identify and discuss several hard-to-decide cases and cases where CHILDES and UD provide different dependency parses. Since only the English databases are annotated with syntactic relations, all labels and discussions here apply only to the English language.

The current version of chatconllu does not handle most of the cases discussed below, because it does not seek to produce gold-standard UD-style dependency parses, but only convert the files to CoNLL-U format, which can then be processed by other dependency parsers, for example those trained on UD treebanks. Therefore, my best effort in designing this mapping goes to transforming existing GR labels to UD labels without changing the dependency structures produced by GRASP in most cases to aid comparison between the two parsing schemes.


% \subsection{Direct Label Translation}

% \paragraph{\texttt{INCROOT}}
% Since CHAT transcripts tries to faithfully transcribe conversational speech, the utterances in CHILDES corpora may not be fully grammatical. This ungrammaticality problem becomes even more apparent when dealing with child speech. For incomplete utterances or sentences missing the main verb, a substitute root \texttt{INCROOT} is given by GRASP. Most of the time, it does not hurt to give it the regular \texttt{ROOT} label directly.

\begin{minipage}{\widefigurewidth}
\begin{dependency}[edge slant=3pt]
	\begin{deptext}[column sep=0.7cm]
	well \& ‡ \& that \& 's \& life \& .\\
	% \textlf{1}|\textlf{0}|BEG \& \textlf{2}|\textlf{1}|BEGP \& \textlf{3}|\textlf{4}|SUBJ \& \textlf{4}|\textlf{0}|ROOT \& \textlf{5}|\textlf{4}|OBJ \& \textlf{6}|\textlf{4}|PUNCT\\
	\end{deptext}
	\depedge{1}{2}{BEGP}
	\depedge{4}{3}{SUBJ}
	\depedge{4}{5}{OBJ}
	\depedge{4}{6}{PUNCT}
	\deproot{1}{BEG}
	\deproot{4}{ROOT}
\end{dependency}
\hfill
\begin{dependency}[edge slant=3pt]
	\begin{deptext}[column sep=0.7cm]
	well \& , \& that \& 's \& life \& .\\
	\end{deptext}
	\depedge{5}{1}{discourse}
	\depedge{5}{2}{punct}
	\depedge{5}{3}{nsubj}
	\depedge{5}{4}{cop}
	\depedge{5}{6}{punct}
	\deproot[edge unit distance=4ex]{5}{root}
\end{dependency}
\end{minipage}
\captionof{figure}{CHILDES vs UD dependency graphs: interjections}\label{fig:intj}

\begin{minipage}{\widefigurewidth}
\begin{dependency}[edge slant=3pt]
	\begin{deptext}[column sep=0.7cm]
	son \& ‡ \& that \& 's \& life \& .\\
	% \textlf{1}|\textlf{0}|BEG \& \textlf{2}|\textlf{1}|BEGP \& \textlf{3}|\textlf{4}|SUBJ \& \textlf{4}|\textlf{0}|ROOT \& \textlf{5}|\textlf{4}|OBJ \& \textlf{6}|\textlf{4}|PUNCT\\
	\end{deptext}
	\depedge{1}{2}{BEGP}
	\depedge{4}{3}{SUBJ}
	\depedge{4}{5}{OBJ}
	\depedge{4}{6}{PUNCT}
	\deproot{1}{BEG}
	\deproot{4}{ROOT}
\end{dependency}
\hfill
\begin{dependency}[edge slant=3pt]
	\begin{deptext}[column sep=0.7cm]
	son \& , \& that \& 's \& life \& .\\
	\end{deptext}
	\depedge{5}{1}{vocative}
	\depedge{5}{2}{punct}
	\depedge{5}{3}{nsubj}
	\depedge{5}{4}{cop}
	\depedge{5}{6}{punct}
	\deproot[edge unit distance=4ex]{5}{root}
\end{dependency}
\end{minipage}
\captionof{figure}{CHILDES vs UD dependency graphs: vocatives}\label{fig:voc}

% \subsection{Difficult Cases}\label{sec:grdifficult}
\subsection{The Multiple-Root Problem}
In UD, every sentence has exactly one root, represented by a single-root dependency tree. The dependency analysis generated by GRASP is, however, not always like that. After processed by GRASP, sentences with non-final punctuations, like after an exclamation, a vocative expression or, close to the end of the utterance, before the tag question, receives multiple roots. Although usually only one \texttt{ROOT} or \texttt{INCROOT} label is assigned, there can be multiple words whose head is marked \textlf{0}, which, in effect, corrupts the single-root tree structure.

This problem occurs because exclamations and vocatives are not considered part of the clause. Tag markers are, however, for the chosen few corpora, not so much of a problem and usually have their head marked to the functional root \sidenote{Although from my observations, tag questions present a different problem of inconsistent parses of the same structure.}(\texttt{ROOT} or \texttt{INCROOT}), like in \Cref{fig:taggr}.

Consider the following sentences:

\pex~ Sentences with non-final punctuations\label{roots}
\a {[\sl Well\/}]$_{\text{\sc intj}}$, that's life.        \hfill {\sl interjection}\label{intj}
\a {[\sl  Son\/}]$_{\text{\sc voc}}$, that's life.         \hfill {\sl vocative}\label{voc}
\a That's life , {[\sl isn't it}]$_{\text{\sc tag}}$?    \hfill {\sl tag question}\label{tagq}
\xe


After running GRASP, sentences like (\ref{intj}) and (\ref{voc}) will be analysed like the left-hand side dependency graph of \Cref{fig:intj} and \Cref{fig:voc}, while UD-style dependeny analyses are shown on the right-hand side.\\

\begin{minipage}{\widefigurewidth}
\begin{dependency}[edge slant=3pt]
	\begin{deptext}[column sep=1cm]
	that \& 's \& life \& „ \& is \& n't \& it \& ?\\
	% \textlf{1}|\textlf{2}|SUBJ \& \textlf{2}|\textlf{0}|ROOT \& \textlf{3}|\textlf{2}|OBJ \& \textlf{4}|\textlf{5}|ENDP \& \textlf{5}|\textlf{2}|END \& \textlf{6}|\textlf{2}|PUNCT\\
	\end{deptext}
	\depedge{2}{1}{SUBJ}
	\depedge{2}{3}{PRED}
	\depedge{5}{4}{ENDP}
	\depedge[edge unit distance=1.8ex]{2}{5}{END}
	\depedge{5}{6}{NEG}
	\depedge[edge unit distance=2.5ex]{5}{7}{PRED}
	\depedge[edge unit distance=1.5ex]{2}{8}{PUNCT}
	\deproot[edge unit distance=4ex]{2}{ROOT}
\end{dependency}
% \captionof{figure}{CHILDES}
\end{minipage}
\captionof{figure}{CHILDES: tag questions}
\label{fig:taggr}


It is clear from the dependency graphs that the CHILDES dependency trees are broken. Therefore, to produce a proper dependency tree from that, one must change the head index of the token identified as the fake root (indicated by the GR label \texttt{BEG}) to that of the sentence's true root.

After the change-head operation is applied, the current token with GR label \texttt{BEG} is connected to the only one root remained. Now the two examples in \Cref{fig:intj} and \Cref{fig:voc} share the same dependency structures. However, dependency relation labels for vocatives and interjections should be different. To make the final decision, chatconllu needs POS tag of the token with \texttt{BEG}. The decision process is outlined below:\\

\begin{itemize}
	\item If the token has a POS tag and it is not \texttt{INTJ}, \texttt{PROPN}, or \texttt{NOUN}, the corresponding \texttt{deprel} \sidenote{Definition for \texttt{parataxis} can be found from \url{https://universaldependencies.org/u/dep/parataxis.html}} label should be \texttt{parataxis} for a loose side-by-side placement of the affected elements.
	\item If the POS tag is \texttt{INTJ}, then the first part of the sentence is considered as an exclamation. The corresponding UD-style label should be \texttt{discourse} \sidenote{Definition for \texttt{discourse} can be found from \url{https://universaldependencies.org/u/dep/discourse.html}}.
	\item If the POS tag is \texttt{PROPN} or \texttt{NOUN}, the first part of the sentence is considered as an expression in vocative. Therefore, the corresponding \texttt{deprel} label should be \texttt{vocative} \sidenote{Definition for \texttt{vocative} can be found from \url{https://universaldependencies.org/u/dep/vocative.html}}.
\end{itemize}

% For instance, the sentence "well, son, that's just life, isn't it ?" will be analysed by GRASP as this:
\subsection{Inconsistent Dependency Parses}

When searching for cases of tag questions, I noticed that the same tag question are parsed differently by GRASP. Right now chatconllu cannot change them into UD-style dependency structures automatically, so the structures, although problematic, are kept as they are. Only the dependency labels are na\"{\i}vely changed to their UD counterparts according to the mapping I created.

\subsection{Predicate Relations}

In GRASP's syntactic analysis scheme, the grammatical relation \texttt{PRED} identifies a nominal or adjectival predicate and the predicate is seen as the argument of its head, which is a verb (\cite{Macwhinney2000}, Part 3, p.63). However, in UD-style dependency annotations, predicate with copula verb constructions favors the predicate as the head, the differences can be seen from the two parses of the same sentence in \Cref{fig:copula}.\\

\begin{minipage}[b]{0.5\linewidth}
\begin{dependency}
	\begin{deptext}[column sep=0.5cm]
	She \& is \& a \& student \& .\\
	% PRON \& AUX \& DET \& NOUN \& PUNCT\\
	\end{deptext}
	\depedge{2}{1}{SUBJ}
	\depedge{2}{4}{PRED}
	\depedge{4}{3}{DET}
	\depedge{2}{5}{PUNCT}
	\deproot{2}{ROOT}
\end{dependency}
\end{minipage}
\hfill
\begin{minipage}[b]{0.5\linewidth}
\begin{dependency}
	\begin{deptext}[column sep=0.5cm]
	She \& is \& a \& student \& .\\
	% PRON \& AUX \& DET \& NOUN \& PUNCT\\
	\end{deptext}
	\depedge{4}{1}{nsubj}
	\depedge{4}{2}{aux}
	\depedge{4}{3}{det}
	\depedge{4}{5}{punct}
	\deproot{4}{root}
\end{dependency}
\end{minipage}\captionof{figure}{CHILDES vs UD: nominal predicate with copula verbs}\label{fig:copula}
\clearpage

However, constructions of predicate with linking verbs, like the example in \Cref{fig:become} with the verb \emph{become}, receive the same dependency structure.\\

\begin{minipage}[b]{0.5\linewidth}
\begin{dependency}
	\begin{deptext}[column sep=0.5cm]
	She \& became \& happier \& .\\
	% PRON \& VERB \& ADJ \& PUNCT\\
	\end{deptext}
	\depedge{2}{1}{SUBJ}
	\depedge{2}{3}{PRED}
	\depedge{2}{5}{PUNCT}
	\deproot{2}{ROOT}
\end{dependency}
\end{minipage}
\begin{minipage}[b]{0.5\linewidth}
\begin{dependency}
	\begin{deptext}[column sep=0.5cm]
	She \& became \& happier \& .\\
	% PRON \& VERB \& ADJ \& PUNCT\\
	\end{deptext}
	\depedge{2}{1}{nsubj}
	\depedge{2}{3}{xcomp}
	\depedge{2}{5}{punct}
	\deproot{2}{root}
\end{dependency}
\end{minipage}
\captionof{figure}{CHILDES vs UD: predicate with linking verb \emph{become}}
\label{fig:become}
\vspace{0.5em}

In fact, although CHILDES' analyses of copula structures are consistent, always making the copula verb the root, UD treats the cases differently. Three types of copula constructions that produce different UD-style parses are identified with comparison to the CHILDES parses in (\cite{liu2021}):
\begin{itemize}
	\item copula structure with statements like the example in \Cref{fig:copula}
	\item copula structures in Wh- and How-questions:\\ The interrogative words are analysed as the subject of the construction in CHILDES, while in UD-style analysis, the predicate becomes the subject.
	\item copula construction in expletives like \emph{there be} expressions:\\
	CHILDES has \emph{there} analysed as the subject, while in UD-style analysis, the predicate in there-be constructions is the subject.
\end{itemize}

This shows that to obtain UD dependency parses require not only a mapping of the labels used for the grammatical relations but also a case-by-case analysis with the annotation conventions of both formalisms in mind, which presents a problem for automatic conversion.\\

Many other cases that are challenging for adapting CHILDES syntactic annotations to UD-style are discussed and implemented in (\cite{liu2021}), which are not handled by chatconllu. They have made the code for conversion public on GitHub \sidenote{\url{https://github.com/zoeyliu18/Parsing_Speech/tree/main/code}}.\\

Apart from that, Liu and Prud’hommeaux observed that depending on the actual meaning of the child, the utterance \emph{man no spoon} can be interpreted in at least three different ways (\cite{liu2021}). Odijk et al. also reported that there are various reasons for a child to produce speech different from the conventional form used by adult speakers. Sometimes the true meanings are concealed by the same transcriptions, which requires \emph{intensive investigation of each phenomenon} by researchers. It seems that for spoken data like child speech, information like context and tones that help us disambiguate meanings are not conveyed by literal transcriptions. Merely looking at utterance texts, it is even hard for a human to decide the real meaning of the utterance. Even if rules are defined for each grammatical relation, incomplete utterances can still produce unintended results in a decision making process. Therefore, instead of coming up case-based rules to convert the syntactic relation labels and structures, chatconllu stores the original label in the token's \texttt{MISC} field and uses a direct mapping to convert the dependency label while keeping the tree structure produced by GRASP, except for sentences with multiple roots.


