\chapter{Introduction} % Main chapter title

\label{Chapter1} % For referencing the chapter elsewhere, use \ref{Chapter1}

% intro
Traditionally, linguistic research has primarily been language-specific, with theories stem from studies on Indo-European languages. In recent decades, there has been much progress in crosslinguistic research, one of which is the \emph{universal dependencies} (UD) project (\cite{nivre2016} and \cite{nivre2020}).

As Joakim Nivre explained in his chapter \emph{Towards a Universal Grammar in Natural Language Processing}, instead of a descriptive and explanatory account of morphosyntax, UD's universal grammar is developed for natural language processing applications(\cite{nivre2015}). Backed by both universal and language-specific tagsets, UD's \emph{CoNLL-U} annotation allows for crosslinguistic corpus sharing. However, most UD treebanks are still based on written text.

Like CoNLL-U, \emph{CHAT} is a standardised annotation scheme. It is part of the \emph{Child Language Data Exchange System} (CHILDES) project (\cite{Macwhinney2000}), and it offers a standard way to transcribe spontaneous speech. Morphosyntactic information can also be added as dependent tiers to the main utterances, and if preferred, it can be generated automatically by running \emph{MOR} (\cite{Macwhinney2000}), \emph{POST} (\cite{parisse2000}) and \emph{GRASP} (\cite{Sagae2004}).

It would be beneficial to use the advantages of both formalisms to compensate each other---CHILDES can provide UD with spontaneous speech treebanks, and UD can offer its tagsets which are widely accepted. To achieve this goal, a two-way converter, chatconllu, is is designed and implemented as a command-line tool in this project.

% explain the choice of the 2 annotations first
The aim of this thesis project is to explore and experiment with format conversion between the above-mentioned annotation schemes via implementing a two-way converter in Python. The scope of analysis is limited to a selected few corpora from the CHILDES database, listed in \Cref{tab:martabdb}. Transcriptions in CHAT format are first converted to CoNLL-U format and then converted back to see which information is lost in the process. With the aid of tools like UDPipe \cite{straka-etal-2016-udpipe}, new information is augmented to the corpora and after chatconllu's conversion, is represented in CHAT-style dependent tiers. To verify the converted and back-converted files, both formats are validated using the tools provided by the two projects---CHILDES and UD.This thesis also serves as a complement to the code on GitHub \sidenote{childes-ud/chatconllu: \url{https://github.com/Mentha7/childes-ud}}.

The thesis is structured as follows: \Cref{Chapter2} gives necessary background information of both annotation schemes. \Cref{Chapter3} introduces the converter and its implementation. \Cref{Chapter4} discusses the mappings from CHILDES to UD. \Cref{Chapter5} explains the validation process and results, and propose future improvements for chatconllu. Finally, with \Cref{Chapter6} I conclude this thesis.

\section{Related Work}

\paragraph{Dependency Parsing for Low-resource Spontaneous Speech} % (fold)
\label{par:zoey}
(\cite{liu2021}) used a low-resource setting, focusing on one child from the Brown Corpus (\cite{brown1973}) and tested the performance of dependency parsers on parent-child conversations. The out-domain parser was trained on written text and performed well on adult speech but not so well on child speech, while the parser trained on a limited amount of in-domain spoken data improved the parsing result of child speech to be comparable with adult speech. As part of their study, they created a semi-automatic adaptation of annotation formats from CHILDES to UD.

\paragraph{The AnnCor CHILDES Treebank}
(\cite{odijk2018anncor}) preprocessed the Dutch CHILDES corpora, augmented them with syntactic annotations using the Alpino parser, and performed partial manual verification. While creating this treebank, they noticed the many disadvantages of annotation discrepancies and argued that annotation guidelines should be explicit and should be strictly followed to produce consistency. They suggest combining human annotations with automatic checks by computer programs. They also pointed out that although the grammar used in child speech is limited since the child is still acquiring the language, assigning adult syntactic structures to their utterances is still valid.
